Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-12-09 .output/_tex/manuscript.tex
work through most of Jan's comments

\raggedbottom
\usepackage{multirow}
\usepackage{centernot}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

Results: The model accurately and precisely describes observed patterns
of lymph node involvement with a compact set of interpretable
parameters. Midline extension of the primary tumor is identified as the
primary risk factor for contralateral involvement, with advanced T-stage
and extensive ipsilateral involvement further increasing risk. The model
highlights key hierarchical dependencies: occult disease in
contralateral LNL III is highly unlikely if upstream LNL II is
clinically negative, and in contralateral LNL IV, occult disease is
exceedingly rare without LNL III involvement.

Conclusions: This model offers an interpretable, probabilistic framework
to inform personalized elective CTV-N volume reduction. For lateralized
tumors that do not cross the midline, it suggests the contralateral neck
may safely be excluded from elective irradiation. For tumors extending
across the midline but with a clinically negative contralateral neck,
elective irradiation could be limited to LNL II, reducing unnecessary
exposure while maintaining oncological safety.}

@@

In treating head and neck squamous cell carcinomas (HNSCC) with
radiotherapy or surgery, both the primary tumor and clinically detected
lymph node metastases are targeted. Current guidelines also include
large portions of the neck in the elective clinical target volume
(CTV-N)
\citep{gregoire_ctbased_2003, gregoire_delineation_2014, gregoire_delineation_2018, eisbruch_intensitymodulated_2002, biau_selection_2019, chao_determination_2002, vorwerk_guidelines_2011, ferlito_elective_2009}
to mitigate the risk of regional recurrences from untreated microscopic
disease undetectable by in-vivo imaging modalities such as computed
tomography (CT), magnetic resonance imaging (MRI), or positron emission
tomography (PET). However, this approach must balance minimizing the
risk of occult disease in the lymphatic drainage region against the
toxicity of unnecessarily irradiating healthy tissue.

These CTV-N guidelines rely on anatomically defined lymph node levels
(LNLs) \citep{gregoire_delineation_2014} and the overall prevalence of
lymph node metastases within these levels. They often recommend
extensive irradiation of both sides of the neck. However, the general
prevalence of metastasis in a given LNL does not correspond to an
individual patient's risk of occult disease in that region, which
depends on their specific state of tumor progression. For example, a
patient with no clinically detectable nodal disease (cN0) who has a
small, clearly lateralized T1 tumor would receive the same contralateral
CTV-N as a patient with significant ipsilateral nodal involvement and an
advanced tumor crossing the mid-sagittal plane: Both patients receive
elective irradiation of the contralateral LNLs II, III, and IVa
\citep{biau_selection_2019}.

To better quantify individualized risk of occult disease, we previously
developed an intuitive probabilistic hidden Markov model (HMM)
\citep{ludwig_hidden_2021, ludwig_modelling_2023}, originally based on a
conceptually similar a Bayesian network model
\citep{pouymayou_bayesian_2019}. However, these models have been limited
to predicting ipsilateral nodal involvement. This work extends the model
to include contralateral risk predictions, enabling more personalized
radiation volume recommendations. By identifying patients with low
contralateral risk, the model could guide reductions in contralateral
CTV-N, thereby decreasing radiation-induced toxicity and improving
quality of life.

@@

\caption{\label{tbl-data-overview}Overview over the five datasets from
four different institutions used to train and evaluate our model. Here,
we briefly characterize the total number of OPSCC patients from the
respective institution, their median age, what proportion received some
form of neck dissection, the N0 portion of patients, what percentage
presented with early T-category, and the prevalence of primary tumor
midline extension. For a much more detailed look at the data, visit
\href{https://lyprox.org}{lyprox.org}.}

@@

\caption{}\label{T_20ee3}\tabularnewline
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Centre Léon Bérard & 325 & 60 & 100\% & 19\% & 69\% & 18\% \\
Inselspital Bern & 74 & 61 & 100\% & 18\% & 66\% & 14\% \\
University Hospital Zurich & 287 & 66 & 26\% & 18\% & 52\% & 31\% \\
Vall d\textquotesingle Hebron Barcelona Hospital & 147 & 58 & 5\% & 21\%
& 34\% & 34\% \\

@@

Pathological involvement is available only for surgically treated
patients and for the levels that were dissected. For non-surgical
patients, involvement status is determined clinically, i.e.~using
imaging. To ensure uniformity in analysis, diagnostic information was
synthesized into a consensus decision for each patient and LNL. This
consensus reflects the most likely state of involvement and accounts for
the sensitivity and specificity of various diagnostic modalities, as
reported in the literature
\citep{debondt_detection_2007, kyzas_18ffluorodeoxyglucose_2008}.

@@

The datasets enable analysis of correlations between LNL involvement and
key risk factors. In figure~\ref{fig-data-strat}, we illustrate the
prevalence of contralateral LNL involvement, stratified by T-category,
the number of ipsilaterally involved LNLs, and whether the tumor extends
across the mid-sagittal plane.

@@

\caption{\label{fig-data-strat}Contralateral involvement stratified by
T-category (top left panel), the number of metastatic LNLs ipsilaterally
(top right panel), and whether the primary tumor extended over the
mid-sagittal line or was clearly lateralized (bottom left panel).
Additionally, in the bottom right, we show that even for lateralized
tumors, T-category and ipsilateral involvement are correlated with
contralateral metastasis by comparing the contralateral involvement
prevalence for selected scenarios that vary in their T-category and
ipsilateral involvement extent.}

@@

The top right panel reveals a positive correlation between ipsilateral
and contralateral metastases. Extensive ipsilateral involvement likely
indicates a longer or faster disease progression. Additionally, it has
been hypothesized that bulky ipsilateral nodal disease may reroute
lymphatic drainage toward the contralateral side, potentially
facilitating contralateral metastasis.

\subsubsection{Interplay of Risk
Factors}\label{interplay-of-risk-factors}

Midline extension, T-category, and ipsilateral involvement are
interrelated risk factors for contralateral metastasis. For instance,
45.6\% of advanced T-category tumors exhibit midline extension compared
to 6.1\% of early T-category tumors. While the higher fraction of
midline extensions in advanced T-category patients partially accounts
for the higher contralateral metastasis rates, T-category itself and
ipsilateral involvement also play an additional role.

The bottom right panel of figure~\ref{fig-data-strat} isolates patients
with lateralized tumors that do not cross the midline, further
demonstrating these relationships. Among early T-category patients with
no ipsilateral nodal involvement (levels I-V), only 1.2\% (1 of 86
patients) show involvement in contralateral level II. This proportion
increases to 8.8\% (24 of 272) if ipsilateral level II is involved, to
15.7\% (14 of 89) if ipsilateral levels II and III are involved, and
further to 22.2\% (12 of 54) for advanced T-category tumors with
ipsilateral levels II and III involved.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Midline Extension:} Tumors extending across the mid-sagittal
  plane should result in a significantly higher probability of
  contralateral metastases.
\item
  \textbf{T-Category and Disease Progression:} Advanced T-category
  should correspond to an increased risk of nodal disease. This can be
  modeled using the expected time of diagnosis, as demonstrated in a
  prior hidden Markov model \citep{ludwig_hidden_2021}.
\item
  \textbf{Ipsilateral Involvement as a Progression Marker:} The extent
  of ipsilateral involvement should provide additional information on
  the time elapsed between disease onset and diagnosis, supplementing
  the insights gained from T-category alone.
\end{enumerate}

@@

This paper builds on the previously developed unilateral model for
ipsilateral lymph node involvement presented in
\citep{ludwig_modelling_2024}. Below, we briefly recap the unilateral
model to introduce the notation required for extending the framework to
a bilateral model, described in section~\ref{sec-ext-to-contra}. For
further details on the ipsilateral model, refer to earlier publications
\citep{ludwig_hidden_2021, ludwig_modelling_2024}.

@@

Here, \(V\) is the number of LNLs the model considers. The values a
LNL's may take on are \(X_v[t] = 0\) (\texttt{False}), meaning the LNL
\(v\) is healthy or free of metastatic disease, or \(X_v[t] = 1\)
(\texttt{True}), corresponding to some form of tumor presence (i.e.,
occult or macroscopic disease). In total, there are \(2^V\) distinct
possible lymphatic involvement patterns, which we enumerate from
\(\boldsymbol{\xi}_0 = \begin{pmatrix} 0 & 0 & \cdots & 0 \end{pmatrix}\)
to
\(\boldsymbol{\xi}_{2^V} = \begin{pmatrix} 1 & 1 & \cdots & 1 \end{pmatrix}\).
Each LNL's state is observed via another binary random variable \(Z_v\)
that describes the clinical involvement of a LNL based on imaging:
\(Z_v = 0\) (\texttt{False}) indicates that the LNL \(v\) is healthy
based on clinical diagnosis, and \(Z_v = 1\) (\texttt{True}) indicates
that LNL \(v\) was classified as involved. \(X_v\) and \(Z_v\) are
connected through the sensitivity and specificity of the diagnositc
modality.

@@

The transition matrix \(\mathbf{A}\) is parameterized using a directed
acyclic graph (DAG) that represents the underlying lymphatic network.
Edges from the primary tumor to an LNL are associated with a probability
\(b_v\) for direct spread to LNL \(v\) during one time step. Arcs from a
LNL \(v\) to a LNL \(r\) are parameterized with the probability rate
\(t_{vr}\) for inter-nodal progression. In this paper, we build on the
DAG shown in figure~\ref{fig-full-graph} which was obtained by
maximizing the model evidence as described in
\citep{ludwig_modelling_2024}.

@@

We can evolve the distribution of all possible hidden states from
\(\mathbf{X}[t=0] = \boldsymbol{\xi}_0\) step by step, by successively
multiplying this vector with the transition matrix \(\mathbf{A}\):
\(\mathbf{X}[t+1] = \mathbf{X}[t] \cdot \mathbf{A}\). For later use, we
define at this point a matrix \(\boldsymbol{\Lambda}\) that collects
these distributions for all considered time-steps:

@@

Where the \(k\)-th row in this matrix corresponds to the probability
distribution over hidden states after \(t=k-1\) time-steps.

@@

\[
P\left( \mathbf{X} \mid \text{T}x = \text{early} \right) = \sum_{t=0}^{t_\text{max}} P \left( \mathbf{X} \mid t \right) \cdot P(t \mid \text{early})
\]

@@

The likelihood function for a patient to present with a diagnosis
\(\boldsymbol{\zeta}_i\) a T-category \(\text{T}x\) tumor can now be
written as:

@@

With \(\left[ \ldots \right]_i\) we denote the \(i\)-th component of the
vector in the square brackets. Note that it is also possible to account
for missing involvement information: If a diagnosis (like fine needle
aspiration (FNA)) is only available for a subset of all LNLs, we can sum
over all those possible complete observed states
\(\boldsymbol{\zeta}_j\) that match the provided diagnosis.

The single-patient likelihood \(\ell\) in
equation~\ref{eq-single-patient-llh} depends on the spread parameters
shown in figure~\ref{fig-full-graph} via the transition matrix
\(\mathbf{A}\) and on the binomial parameters \(p_{\text{T}x}\) via
time-priors. In this work, we will only differentiate between ``early''
(T1 \& T2) and ``advanced'' (T3 \& T4) T-categories. Therefore, the
parameter space of the unilateral model is:

@@

A straightforward approach to modeling contralateral lymphatic spread
would be to use two independent unilateral models, as described in
section~\ref{sec-unilateral}, with shared parameters such as the
distribution of diagnosis times or spread between LNLs (\(t_{vr}\)).
However, this method would fail to capture the correlation between
ipsilateral and contralateral involvement discussed in
section~\ref{sec-data-strat}, particularly the observed increase in
contralateral involvement with greater severity of ipsilateral spread.

Thus, we extend the formalism in section~\ref{sec-unilateral} in such a
way that the model's ipsi- and contralateral side evolve synchronously
over time. To achieve that, we start by writing down the posterior
distribution of involvement in analogy to
equation~\ref{eq-uni-bayes-law}, which is now a joint probability of an
involvement \(\mathbf{X}^\text{i}\) ipsilaterally \emph{and} an
involvement \(\mathbf{X}^\text{c}\) contralaterally, given a diagnosis
of the ipsilateral LNLs \(\mathbf{Z}^\text{i}\) and of the contralateral
ones \(\mathbf{Z}^\text{c}\):

@@

The matrices \(\boldsymbol{\Lambda}\text{i}\) and
\(\boldsymbol{\Lambda}\text{c}\) could, in principle, be parameterized
with entirely separate parameters, allowing ipsilateral and
contralateral spread rates to differ completely. However, we simplify
the parameter space by sharing parameters between the two sides, based
on the following three assumptions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Shared Graph Structure}: Both ipsilateral and contralateral
  spread are described by the same graph shown in
  figure~\ref{fig-full-graph}.
\item
  \textbf{Symmetric Spread Among LNLs}: The spread among LNLs is assumed
  to be the same on both sides, reflecting the symmetric structure of
  the lymphatic system. Consequently, the spread rates between nodes
  should also be symmetric. This is formalized as:\\
  \begin{equation}\phantomsection\label{eq-symmetries}{
    \begin{aligned}
    b_v^\text{c} &\neq b_v^\text{i} \\
    t_{rv}^\text{c} &= t_{rv}^\text{i}
    \end{aligned}
    }\end{equation} for all \(v \leq V\) and
  \(r \in \operatorname{pa}(v)\).
\item
  \textbf{Asymmetric Spread from Tumor}: Tumor spread to the
  contralateral side varies depending on whether the tumor crosses the
  mid-sagittal plane. When the tumor extends over the midline, its
  spread rate to the contralateral LNLs should be larger than if the
  tumor were lateralized, but still smaller than the tumor's spread to
  the ipsilateral side. To model this, we introduce a mixing parameter
  \(\alpha \in [0,1]\), representing the ``degree of asymmetry''
  effectively mixing these two extremes:\\
  \begin{equation}\phantomsection\label{eq-mixing}{
    b_v^{\text{c},\epsilon=\texttt{True}} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\epsilon=\texttt{False}}
    }\end{equation} This parameterization results in three sets of rates
  for tumor spread to the LNLs: (1) the spread to ipsilateral LNLs
  \(b^\text{i}_v\), (2) the spread to contralateral LNLs as long as the
  tumor is lateralized \(b_v^{\text{c},\epsilon=\texttt{False}}\), (3)
  the spread to contralateral LNLs when the tumor crosses the midline
  \(b_v^{\text{c},\epsilon=\texttt{True}}\). Despite this additional
  complexity, these rates collectively account for only
  \(2 \cdot 2^V + 1\) parameters due to their coupling via \(\alpha\).
\end{enumerate}

The parameter space now becomes:

@@

Most tumors crossing the midline at the time of diagnosis likely began
as lateralized tumors that grew over the midline at a later stage. As a
result, the transition matrix
\(\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}\) applies only to a
subset of time-steps.

To account for this, we model the tumor's extension over the
mid-sagittal line as a binary random variable. A tumor starts as
lateralized, with a finite probability \(p_\epsilon\) at each time step
of crossing the midline. The overall probabilities of a patient having a
clearly lateralized tumor or one extending over the mid-sagittal line
after \(t\) time steps are then given by

@@

Using this, it is straightforward to write down the matrix of state
distributions for all time-steps, as in equation~\ref{eq-lambda-matrix}
covering the contralateral hidden state evolution:

@@

where we used the transition matrix
\(\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}\) that depends on the
base spread parameters \(b_v^{\text{c},\epsilon=\texttt{False}}\).

The case of midline extension is more complex: we already marginalize
over the exact time step when the tumor grows over the mid-sagittal
line. However, at the point of crossing, the contralateral transition
matrix must switch to the increased spread rates,
\(b_v^{\text{c}, \epsilon=\texttt{True}}\), as defined by the linear
mixing in equation~\ref{eq-mixing}. To correctly perform this
marginalization, we iteratively construct the joint distribution
\(P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)\).

@@

\subsection{Involvement Data Consensus}\label{sec-methods-consensus}

>>>>

\raggedbottom
\usepackage{multirow}
\usepackage{centernot}
\usepackage{mathtools}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

Results: The model accurately and precisely describes observed patterns
of lymph node involvement with a compact set of interpretable
parameters. Midline extension of the primary tumor is identified as the
primary risk factor for contralateral involvement, with advanced T-stage
and extensive ipsilateral involvement further increasing risk. Occult
disease in contralateral LNL III is highly unlikely if upstream LNL II
is clinically negative, and in contralateral LNL IV, occult disease is
exceedingly rare without LNL III involvement.

Conclusions: This model offers an interpretable, probabilistic framework
to inform personalized elective CTV-N volume reduction. For lateralized
tumors that do not cross the midline, it suggests the contralateral neck
may safely be excluded from elective irradiation. For tumors extending
across the midline but with a clinically negative contralateral neck,
elective irradiation could be limited to LNL II, reducing unnecessary
exposure of normal tissue while maintaining regional tumor control.}

@@

In head and neck squamous cell carcinomas (HNSCC) treatments with
radiotherapy or surgery, both the primary tumor and clinically detected
lymph node metastases are targeted. In addition, current guidelines
include large portions of the neck in the elective clinical target
volume (CTV-N)
\citep{gregoire_ctbased_2003, gregoire_delineation_2014, gregoire_delineation_2018, eisbruch_intensitymodulated_2002, biau_selection_2019, chao_determination_2002, vorwerk_guidelines_2011, ferlito_elective_2009}
to mitigate the risk of regional recurrences from untreated microscopic
disease undetectable by in-vivo imaging modalities such as computed
tomography (CT), magnetic resonance imaging (MRI), or positron emission
tomography (PET). However, this approach must balance minimizing the
risk of occult disease in the lymphatic drainage region against the
toxicity of unnecessarily irradiating healthy tissue.

These CTV-N guidelines rely on anatomically defined lymph node levels
(LNLs) \citep{gregoire_delineation_2014} and the overall prevalence of
lymph node metastases within these levels. They often recommend
extensive irradiation of both sides of the neck. However, the general
prevalence of metastasis in a given LNL does not correspond to an
individual patient's risk of occult disease in that region, which
depends on their specific state of tumor progression. For example, a
patient with no clinically detectable nodal disease (cN0) who has a
small, clearly lateralized T1 tumor would receive the same contralateral
CTV-N as a patient with significant ipsilateral nodal involvement and an
advanced tumor crossing the mid-sagittal plane. Both patients receive
elective irradiation of the contralateral LNLs II, III, and IVa
\citep{biau_selection_2019}.

To better quantify individualized risk of occult disease, we previously
developed an intuitive probabilistic hidden Markov model (HMM)
\citep{ludwig_hidden_2021, ludwig_modelling_2023}, originally based on a
conceptually similar a Bayesian network model
\citep{pouymayou_bayesian_2019}. However, these models have been limited
to predicting ipsilateral nodal involvement. This work extends the model
to include contralateral risk predictions, enabling more personalized
radiation volume recommendations for the contralateral neck. By
identifying patients with low contralateral risk, the model could guide
reductions in the contralateral CTV-N, thereby decreasing
radiation-induced toxicity and improving quality of life.

@@

\caption{\label{tbl-data-overview}Overview over the four datasets from
four different institutions used to train and evaluate our model. Here,
we briefly characterize the total number of OPSCC patients from the
respective institution, their median age, what proportion received neck
dissection, the N0 portion of patients, what percentage presented with
early T-category (T1/T2), and the prevalence of primary tumor midline
extension. For a much more detailed look at the data, visit
\href{https://lyprox.org}{lyprox.org}.}

@@

\caption{}\label{T_fbab8}\tabularnewline
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Centre Léon Bérard & 325 & 60 & 100\% & 19\% & 69\% & 18\% \\
Inselspital Bern & 74 & 61 & 100\% & 18\% & 66\% & 14\% \\
University Hospital Zurich & 287 & 66 & 26\% & 18\% & 52\% & 31\% \\
Vall d\textquotesingle Hebron Barcelona Hospital & 147 & 58 & 5\% & 21\%
& 34\% & 34\% \\

@@

Pathological involvement is available only for surgically treated
patients and for the levels that were dissected. For non-surgical
patients, involvement status is determined clinically, i.e.~using
imaging. For this work, diagnostic information was synthesized into a
consensus decision for each patient and LNL. This consensus reflects the
most likely state of involvement and accounts for the sensitivity and
specificity of various diagnostic modalities, as reported in the
literature
\citep{debondt_detection_2007, kyzas_18ffluorodeoxyglucose_2008}.

@@

The datasets enable analysis of correlations between contralateral LNL
involvement and key risk factors. In figure~\ref{fig-data-strat}, we
illustrate the prevalence of contralateral LNL involvement, stratified
by T-category, the number of ipsilaterally involved LNLs, and whether
the tumor extends across the mid-sagittal plane.

@@

\caption{\label{fig-data-strat}Contralateral involvement stratified by
T-category (top left panel), the number of metastatic LNLs ipsilaterally
(top right panel), and whether the primary tumor extended over the
mid-sagittal plane or was clearly lateralized (bottom left panel). In
the bottom right panel, we consider lateralized tumors only, and compare
the contralateral involvement prevalence for selected scenarios that
vary in their T-category and ipsilateral involvement extent.}

@@

The top right panel reveals a positive correlation between ipsilateral
and contralateral metastases. Extensive ipsilateral involvement likely
indicates a longer or faster disease progression. Additionally, it has
been hypothesized that bulky ipsilateral nodal disease may reroute
lymphatic drainage toward the contralateral side, potentially increasing
the probability of contralateral metastasis.

\subsubsection{Correlation of Risk
Factors}\label{correlation-of-risk-factors}

Midline extension, T-category, and ipsilateral involvement are
interrelated risk factors for contralateral metastasis. For instance,
45.6\% of advanced T-category tumors exhibit midline extension compared
to 6.1\% of early T-category tumors. While the higher fraction of
midline extensions in advanced T-category patients partially explains
the higher contralateral metastasis rates, T-category itself and
ipsilateral involvement also play an additional role.

The bottom right panel of figure~\ref{fig-data-strat} considers only
patients with lateralized tumors that do not cross the midline. Among
early T-category patients with no ipsilateral nodal involvement (levels
I-V), only 1.2\% (1 of 86 patients) show involvement in contralateral
level II. This proportion increases to 8.8\% (24 of 272) if ipsilateral
level II is involved, to 15.7\% (14 of 89) if ipsilateral levels II and
III are involved, and further to 22.2\% (12 of 54) for advanced
T-category tumors with ipsilateral levels II and III involved.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Midline Extension:} Tumors extending across the mid-sagittal
  plane should result in a significantly higher probability of
  contralateral metastases.
\item
  \textbf{T-Category:} Advanced T-category should correspond to an
  increased risk of nodal disease. In the hidden Markov model this can
  be modeled using the expected time of diagnosis, as demonstrated
  previously for the ipsilateral model \citep{ludwig_hidden_2021}.
\item
  \textbf{Ipsilateral Involvement:} The model should be able to capture
  the correlation between the extent of ipsi- and contralateral
  involvement. I.e., a more severe ipsilateral involvement should
  indicate a higher risk for contralateral metastases.
\end{enumerate}

@@

This paper builds on the previously developed unilateral model for
ipsilateral lymph node involvement presented in
\citep{ludwig_modelling_2024}. Below, we briefly recap the unilateral
model to introduce the notation required for extending the framework to
a bilateral model, described in section~\ref{sec-ext-to-contra}. For
further details on the ipsilateral model, we refer to earlier
publications \citep{ludwig_hidden_2021, ludwig_modelling_2024}.

@@

Here, \(V\) is the number of LNLs the model considers. The values a LNL
may take on are \(X_v[t] = 0\) (\texttt{False}), meaning the LNL \(v\)
is free of metastatic disease, or \(X_v[t] = 1\) (\texttt{True}),
corresponding to the presence of clinically detected metastases (i.e.,
occult or macroscopic disease). In total, there are \(2^V\) distinct
possible lymphatic involvement patterns, which we enumerate from
\(\boldsymbol{\xi}_0 = \begin{pmatrix} 0 & 0 & \cdots & 0 \end{pmatrix}\)
to
\(\boldsymbol{\xi}_{2^V} = \begin{pmatrix} 1 & 1 & \cdots & 1 \end{pmatrix}\).
Each LNL's state is observed via another binary random variable \(Z_v\)
that describes the clinical involvement of a LNL based on imaging:
\(Z_v = 0\) (\texttt{False}) indicates that the LNL \(v\) is healthy
based on clinical diagnosis, and \(Z_v = 1\) (\texttt{True}) indicates
that LNL \(v\) was classified as involved. \(X_v\) and \(Z_v\) are
connected through the sensitivity and specificity of the diagnositc
modality.

@@

The transition matrix \(\mathbf{A}\) is parameterized using a directed
acyclic graph (DAG) that represents the underlying lymphatic network.
Edges from the primary tumor to an LNL are associated with a probability
\(b_v\) for direct spread to LNL \(v\) during one time step. Arcs from a
LNL \(v\) to a LNL \(r\) are parameterized with the probability rate
\(t_{vr}\) representing the probability of spread to an LNL \(r\) that
receives efferent lymphatic spread from LNL \(v\). In this paper, we
build on the DAG shown in figure~\ref{fig-full-graph} which was obtained
by maximizing the model evidence as described in
\citep{ludwig_modelling_2024}.

@@

Let us now consider the probability distribution over all possible
hidden states \(\mathbf{X}[t]\) at time \(t\). We can get to this
distribution by evolving the healthy starting state
\(\mathbf{X}[t=0] = \boldsymbol{\xi}_0\) at time \(t=0\) step by step,
by successively multiplying this vector with the transition matrix
\(\mathbf{A}\): \(\mathbf{X}[t+1] = \mathbf{X}[t] \cdot \mathbf{A}\).
For later use, we define at this point a matrix \(\boldsymbol{\Lambda}\)
that collects these distributions for all considered times:

@@

where the \(k\)-th row in this matrix corresponds to the probability
distribution over hidden states after \(t=k-1\) time-steps.

@@

The probability distribution over \(\mathbf{X}\) for the example of an
early T-category patient is given by

\[
P\left( \mathbf{X} \mid \text{T}x = \text{early} \right) = \sum_{t=0}^{t_\text{max}} P \left( \mathbf{X} \mid t \right) \cdot P(t \mid \text{T}x = \text{early})
\]

@@

The probabiltiy for a patient to present with a diagnosis
\(\mathbf{Z} = \boldsymbol{\zeta}_i\) a T-category \(\text{T}x\) tumor
can now be written as:

@@

here, \(\left[ \ldots \right]_i\) we denote the \(i\)-th component of
the vector in the square brackets. Note that it is also possible to
account for missing involvement information: If a diagnosis (like fine
needle aspiration (FNA)) is only available for a subset of all LNLs, we
can sum over all those possible complete observed states
\(\boldsymbol{\zeta}_j\) that match the provided diagnosis.

The term above represents a single patient's contribution to the overall
likelihood function that is a product of such terms for each patient.
This single-patient likelihood \(\ell\) in
equation~\ref{eq-single-patient-llh} depends on the spread parameters
shown in figure~\ref{fig-full-graph} via the transition matrix
\(\mathbf{A}\) and on the binomial parameters \(p_{\text{T}x}\) via
time-priors. In this work, we will only differentiate between ``early''
(T1 \& T2) and ``advanced'' (T3 \& T4) T-categories. Therefore, the
parameter space of the unilateral model is:

@@

A straightforward approach to modeling contralateral lymphatic spread
would be to use two independent unilateral models, as described in
section~\ref{sec-unilateral}, possibly with shared parameters such as
the distribution of diagnosis times or spread between LNLs (\(t_{vr}\)).
However, this method would fail to capture the correlation between
ipsilateral and contralateral involvement discussed in
section~\ref{sec-data-strat}, particularly the observed increase in
contralateral involvement with greater severity of ipsilateral spread.

Thus, we extend the formalism in section~\ref{sec-unilateral} in such a
way that the model's ipsi- and contralateral side evolve synchronously
over time. To achieve that, we start by writing down the posterior
distribution of involvement, which is now a joint probability of an
involvement \(\mathbf{X}^\text{i}\) ipsilaterally \emph{and} an
involvement \(\mathbf{X}^\text{c}\) contralaterally, given a diagnosis
of the ipsilateral LNLs \(\mathbf{Z}^\text{i}\) and of the contralateral
ones \(\mathbf{Z}^\text{c}\):

@@

The matrices \(\boldsymbol{\Lambda}_\text{i}\) and
\(\boldsymbol{\Lambda}_\text{c}\) could, in principle, be parameterized
with entirely separate parameters, allowing ipsilateral and
contralateral spread rates to differ substantially. However, we simplify
the parameter space by sharing parameters between the two sides, based
on the following three assumptions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Shared Graph Structure}: Both ipsilateral and contralateral
  spread are described by the same graph shown in
  figure~\ref{fig-full-graph}.
\item
  \textbf{Symmetric Spread Among LNLs}: The spread among LNLs is assumed
  to be the same on both sides, reflecting the symmetric structure of
  the lymphatic system. Consequently, the spread rates between nodes
  should also be symmetric. This is formalized as:\\
  \begin{equation}\phantomsection\label{eq-symmetries}{
    t_{rv}^\text{c} = t_{rv}^\text{i}
    }\end{equation} for all \(v \leq V\) and
  \(r \in \operatorname{pa}(v)\) being all nodes that spread to \(v\).
\item
  \textbf{Asymmetric Spread from Tumor}: The probabilities of direct
  spread from the primary tumor are clearly different for the ipsi- and
  contralateral neck. In addition, tumor spread to the contralateral
  side varies depending on whether the tumor crosses the mid-sagittal
  plane. This would result in three sets of rates for tumor spread to
  the LNLs: (1) the spread to ipsilateral LNLs \(b^\text{i}_v\), (2) the
  spread to contralateral LNLs as long as the tumor is lateralized
  \(b_v^{\text{c},\epsilon=\texttt{False}}\), (3) the spread to
  contralateral LNLs when the tumor crosses the midline
  \(b_v^{\text{c},\epsilon=\texttt{True}}\). In this work, however, we
  chose to define the latter set as a linear mix of ipsilateral tumor
  spread and contralateral spread in case of a clearly lateralized
  tumor. Thus, using \(\alpha \in [0,1]\) as this mixing parameter, we
  have \begin{equation}\phantomsection\label{eq-mixing}{
    b_v^{\text{c},\epsilon=\texttt{True}} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\epsilon=\texttt{False}}
    }\end{equation} In this way, the tumor's midline extension causes
  the contralateral spread to become more like the spread to the
  ipsilateral side.
\end{enumerate}

The full parameter space of this model is now:

@@

Most tumors crossing the midline at the time of diagnosis likely began
as lateralized tumors that grew over the midline at a later point in
time. As a result, the transition matrix
\(\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}\) applies only to a
subset of time-steps.

To account for this, we model the tumor's extension over the
mid-sagittal plane as an additional binary random variable \(\epsilon\).
A tumor starts as lateralized, with a finite probability \(p_\epsilon\)
at each time step of crossing the midline. The overall probabilities of
a patient having a clearly lateralized tumor or one extending over the
mid-sagittal plane after \(t\) time steps are then given by

@@

Using this, it is straightforward to write down the matrix of state
distributions for all time-steps, as in equation~\ref{eq-lambda-matrix},
covering the joint distribution over the contralateral hidden state and
the midline extension:

@@

here, we used the transition matrix
\(\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}\) that depends on the
base spread parameters \(b_v^{\text{c},\epsilon=\texttt{False}}\).

The case of midline extension is more complex: we already marginalize
over the exact time step when the tumor grows over the mid-sagittal
plane. However, at the point of crossing, the contralateral transition
matrix must switch to the increased spread rates,
\(b_v^{\text{c}, \epsilon=\texttt{True}}\), as defined by the linear
mixing in equation~\ref{eq-mixing}. To correctly perform this
marginalization, we iteratively construct the joint distribution
\(P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)\).

@@

\subsection{Training Data}\label{sec-training-data}
